Tidy data structure to support exploration and modeling of temporal data

### 2019-01-06 main.Rmd
updated flights case study

## Time-based pipeline {#sec:timepipeline}

@@

As a special case of data frame, a tsibble pipes into the grammar of graphics straight way, making most use of this powerful graphical system. It should be easy to create and extend some specialist plotting methods based on tsibble structure, such as autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

A closer examination of New York airports will give an indication about how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for EWR, JFK and LGA airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are framed as grey lines in Figure \@ref(fig:sel-monthly-plot). LGA fluctuates a lot compared to the other two. What superimposes on these lines is two-month moving averages so that a temporal trend is more visible. The number of days for each month is variable. Moving averages for two months call for computing weighted mean. But this can also be accomplished using a pair of commonly used verbs--`nest()` and `unnest()` to handle list-columns, without worrying weights specification. The sliding with large window size smoothes out the fluctuations and gives a stable trend around 25% over the year.

@@

(ref:sel-monthly-plot-cap) Daily delayed percentages for departure with two-month moving averages overlaid at five international airports. There are many fluctuations observed at LGA airport. The estimates of temporal trend are around 25% across three airports, highlighting relatively less delay in Fall.

@@

Sensors have been put up to collect data for the project of smart city across major cities in Australia. One of the trials is monitoring households' electricity usage through installed smart meters in the area of New Castle over 2010--2014 [@smart-meter]. Year 2013 has been sliced to examine temporal patterns of customer's energy consumption with **tsibble** in this paper. Half-hourly general supply in kwH have been recored for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Customer's demographic data provides explanatory variables other than time in a different data table. Two data tables might be joined to explore different sources that contribute to daily electricity use when needed.

@@

@@

(ref:calendar-plot-cap) Half-hourly average electricity use.

>>>>

## Time-based transformation {#sec:timepipeline}

@@

## Time series visualization and models {#sec:ts-models}

As a special case of the data frame, a tsibble pipes into the grammar of graphics straight way, making most use of this powerful graphical system. It should be easy to create and extend some specialized time-related plotting methods based on tsibble structure, such as autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

A closer examination of some big airports across the US will give an indication about how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for Houston (IAH), New York (JFK), Kalaoa (Hawaii), Los Angeles (LAX) and Seatle (SEA) airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are framed as grey lines in Figure \@ref(fig:sel-monthly-plot). Winter months tend to fluctuate a lot compared to the summer across all the airports. What superimposes on these lines is two-month moving averages so that a temporal trend is more visible. The number of days for each month is variable. Moving averages for two months call for computing weighted mean. But this can also be accomplished using a pair of commonly used verbs--`nest()` and `unnest()` to handle list-columns, without worrying weights specification. The sliding operation with large window size smoothes out the fluctuations and gives a stable trend around 25% over the year. The LAX airport has seen a gradual decline in delays over the year, whereas the SEA airport has a steady control of delays over time. The IAH and JFK airports have undergone more delays in the middle of year, while the KOA has the inverse pattern with higher delay percentage in both ends of the year.

@@

(ref:sel-monthly-plot-cap) Daily delayed percentages for departure with two-month moving averages overlaid at five international airports. There are least fluctuations and relatively fewer delays observed at KOA airport. The estimates of temporal trend are around 25% across the other four airports, but highlight different time periods of severe delays.

@@

Sensors have been put up to collect data for the project of smart city across major cities in Australia. One of the trials is monitoring households' electricity usage through installed smart meters in the area of Newcastle over 2010--2014 [@smart-meter]. Year 2013 has been sliced to examine temporal patterns of customer's energy consumption with **tsibble** in this paper. Half-hourly general supply in kwH have been recored for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Customer's demographic data provides explanatory variables other than time in a different data table. Two data tables might be joined to explore different sources that contribute to daily electricity use when needed.

@@

Aggregation over all the individuals helps to grasp a big picture about if the behaviors are affected by the time.

@@

(ref:calendar-plot-cap) Half-hourly average electricity use across all customers in the region, organized in the calendar format. Distinctive patterns between summer and winter months occur to daily residential energy consumption. Air conditioning drives the peak in the late afternoon in January and December, however, peaks in the morning and evening are more pronounced in the winter months due to heating use.

```{r count-gaps, fig.height = 7, fig.cap = "(ref:count-gaps-cap)"}
```
